---
title: "Cyclistic bike-share analysis case study!"
author: "Ameyaw Ahmed Analysis"
date: '2022-07-15'
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Case Study: Does Cyclistic future successs depends on annual membership riders?
The main goal of this document is to consolidate the downloaded Divvy data into one data frame and then conduct a simple analysis which would help the marketing team to understand **how  casual riders usage of Divvy bikes differ from annual membership riders** by finding  insights that will help that could be turned into recommendation to help the marketing team design a new marketing strategy to convert casual riders into annual members. 

# Introduction
This exploratory analysis case study is towards Capstome project requirement for Google Data Analytics Professional Certificate. The case study involves a bikeshare company's data of its customer's trip details over a **13 month period (Jul 2021 - Jul 2022)**. The data has been made available by Motivate International Inc. under this [license](https://ride.divvybikes.com/data-license-agreement).

The analysis will follow the 6 phases of the Data Analysis process: Ask, Prepare, Process, Analyze, and Act. A brief explanation of these processes:

# Ask
- Ask effective questions
- Define the scope of the analysis
- Define what success looks like

# Prepare
- Verify data integrity
- Check data credibility and reliability
- Check data types
- Merge datasets

# Process
- Clean, Remove and Transform data
- Document cleaning processes and results

# Analyze
- Identify patterns
- Draw conclusions
- Make predictions

# Share
- Create effective visuals
- Create a story for data
- Share insights to stakeholders

# Act
- Give recommendations based on insights
- Solve problems
- Create something new

# 1. Ask
**Scenario**: The cyclistic marketing team needs to design strategies aimed atconverting casual riders into annual members. Howver the marketing team needs to under how the casual riders and annual members differ.

# Stakeholders
- Director of marketing
-Cyclistic executive team

# Objective
The main goal is therefore to analysis the 13-month period data to extract some insight on how the two main customer sgegment use the Cyclistic bikes differently.

# Deliverables
- Provide insights on how annual riders and casual riders use the Cyclistic bikes differently
- Support the findings with effective visualizations
- Based on the insights found, give recommendations on how to convert casual riders to member riders.

# 2. Prepare and Process

######  Source of data
The data has been made available by Motivate International Inc. under this [license](https://www.divvybikes.com/data-license-agreement). The data is a monthly data which begins at July 11th, 2021 and ends at July 14th 2021. The data can be found [here](https://divvy-tripdata.s3.amazonaws.com/index.html). The data consist of a toal of **13 CSV files** starting from July 2021 to July 2022. Each of the files contain data of the monthly rides by customers of Cyclistic. The combined data is over 6 million observations hence doing data cleaning and  wrangling in excel will be slow thats why I have chosen to conduct all the analysis process in R. 

## Load Libraries
```{r message=FALSE, warning=FALSE}
library(tidyverse)
library(janitor)
library(here)
library(lubridate)
library(hms)
library(anytime)
library(scales)
library(zoo)
library(ggcharts)
```

## Impoting files
```{r include=FALSE}
Jul_firstyr <- read_csv(here::here("Dataset/Jul_2021_tripdata.csv"))
Aug_firstyr <- read_csv(here::here("Dataset/Aug_2021_tripdata.csv"))
Sep_firstyr <- read_csv(here::here("Dataset/Sep_2021_tripdata.csv"))
Oct_firstyr <- read.csv(here::here("Dataset/Oct_2021_tripdata.csv"))
Nov_firstyr <- read_csv(here::here("Dataset/Nov_2021_tripdata.csv"))
Dec_firstyr <- read_csv(here::here("Dataset/Dec_2021_tripdata.csv"))
Jan_secyr <- read_csv(here::here("Dataset/Jan_2022_tripdata.csv"))
Feb_secyr <- read_csv(here::here("Dataset/Feb_2022_tripdata.csv"))
Mar_secyr <- read_csv(here::here("Dataset/Mar_2022_tripdata.csv"))
Apr_secyr <- read.csv(here::here("Dataset/Apr_2022_tripdata.csv"))
May_secyr <- read_csv(here::here("Dataset/May_2022_tripdata.csv"))
Jun_secyr <- read_csv(here::here("Dataset/Jun_2022_tripdata.csv"))
Jul_secyr <- read_csv(here::here("Dataset/Jul_2022_tripdata.csv"))
```

### Comparaing datatype of the monthly data to further combine them
```{r}
compare_df_cols(
  Apr_secyr,
  Dec_firstyr,
  Aug_firstyr,
  Feb_secyr,
  Jan_secyr,
  Jul_firstyr,
  Jul_secyr,
  Jun_secyr,
  Mar_secyr,
  May_secyr,
  Nov_firstyr,
  Oct_firstyr,
  Sep_firstyr
)
```

Observing from the structure of the data, there are inconsistencies in the datatype, the **started_at** and **ended_at** of **Apr_secyr* and **Oct_firstyr**  are stored as characters instead of dates hence I convert these variables into the right type of datatype.

```{r}
Apr_secyr$started_at <- ymd_hms(Apr_secyr$started_at)
Apr_secyr$ended_at <- ymd_hms(Apr_secyr$ended_at)
Oct_firstyr$started_at <- ymd_hms(Oct_firstyr$started_at)
Oct_firstyr$ended_at <- ymd_hms(Oct_firstyr$ended_at)
```

Now that all datatype are consistent lets combine all dataframe into a single dataframe
### Combining all the monthly dataframe into one dataframe
```{r}
bike_rides_df <-
  rbind(
    Apr_secyr,
    Aug_firstyr,
    Dec_firstyr,
    Feb_secyr,
    Jan_secyr,
    Jul_firstyr,
    Jul_secyr,
    Jun_secyr,
    Mar_secyr,
    May_secyr,
    Nov_firstyr,
    Oct_firstyr,
    Sep_firstyr
  )
```

Since we don't need the individual dataframe, lets remove them to have a clean environment to work with.
### Reomve individual dataframe 
```{r}
rm(
  Apr_secyr,
  Dec_firstyr,
  Aug_firstyr,
  Feb_secyr,
  Jan_secyr,
  Jul_firstyr,
  Jul_secyr,
  Jun_secyr,
  Mar_secyr,
  May_secyr,
  Nov_firstyr,
  Oct_firstyr,
  Sep_firstyr
)
```


### viewing the data combined data
```{r}
glimpse(bike_rides_df)
head(bike_rides_df)
str(bike_rides_df)
```


### To compute the length of each ride, I need to exxtract start and end time from the started_at and ended_at colimns. I extracted hours, mins and secs from the started_at and ended_at columns and add new columns which will be start_time and end_time

```{r}
bike_rides_df$start_time <-
  hms::as_hms(bike_rides_df$started_at)
bike_rides_df$end_time <-
  hms::as_hms(bike_rides_df$ended_at)
```

### Extraction of start date and end date
```{r}
bike_rides_df$start_date <-
  as_date(bike_rides_df$started_at)
bike_rides_df$end_date <-
  as_date(bike_rides_df$ended_at)
```

```{r}
bike_rides_df$start_hour <- 
  strptime(bike_rides_df$start_time, "%H") %>%
  hour()
```

### Extracting months from the date and converting them to month names
```{r}
bike_rides_df$month <-
  as.yearmon(bike_rides_df$start_date)
```


### Some of the variables are irrelevant in answering the questions I aim to answer. Hence I will remove such coilunns

```{r}
bike_rides_df <- bike_rides_df %>%
  select(
    rideable_type,
    started_at,
    ended_at,
    start_station_name,
    end_station_name,
    member_casual,
    start_time,
    end_time,
    start_date,
    month,
    start_hour
  ) %>%
  rename(
    "bike_type" = rideable_type,
    "customer_type" = member_casual,
    "ride_date" = start_date
  )
```

### Extract day of the week from date the trip started
```{r}
bike_rides_df$week_day <- weekdays(bike_rides_df$ride_date)
```


### Create trip length from the started_at and ended_at

```{r}
options(scipen = 999)
bike_rides_df$trip_duration <-
  as.double(difftime(bike_rides_df$end_time, bike_rides_df$start_time)) / 60
```

To avoid errors in our data visualization and error in the analysis, we need to check if there is and trip duration that was less than 0. If there is, we will exclude them. These trips with less than 0 trip duration were / could be as a result of the test trip conducted by the company hence we will exclude them. The trips with their start station containing **"test"** in their names are the test trip the company made.

# Examining if there is a trip duration that was less than 0
```{r}
bike_rides_df %>%
  select(trip_duration) %>%
  filter(trip_duration < 0) %>%
  count()

# count how many trips were test trips made by the company
grep("test", bike_rides_df$start_station_name, value = TRUE)

grep("Test", bike_rides_df$start_station_name, value = TRUE)

grep("TEST", bike_rides_df$start_station_name, value = TRUE)
```

### Removing all trip duration that were less than 0
```{r}
bike_rides_df_new <- bike_rides_df[!(bike_rides_df$trip_duration < 0), ]

# Check if the test trip and station is removed
grep("Test", bike_rides_df_new$start_station_name, value = TRUE)

rm(bike_rides_df)
```

### Removing test trip
```{r}
bike_rides_df_new <-
  bike_rides_df_new[!grepl("Test", bike_rides_df_new$start_station_name), ]
```


### view final dataframe before analysis  and give levels to customer type and bike type
```{r}
bike_rides_df_new$customer_type <-
  factor(
    bike_rides_df_new$customer_type,
    levels = c("member", "casual")
  )
bike_rides_df_new$bike_type <- 
  factor(
    bike_type, 
    levels = c("classic_bike", "electric_bike", "docked_bike")
  )
glimpse(bike_rides_df_new)
```

# Analyze & Share

Our data frame is now ready to uncover some insights on how casual riders use Cyclistic bikes differently from member riders.
Lets first of all take a statistical summary of the trip duration

```{r warning=FALSE}
attach(bike_rides_df_new)
summary(trip_duration)
```

# Agreggate summary by customer type
```{r}
bike_rides_df_new %>%
  select(trip_duration, customer_type) %>%
  group_by(customer_type) %>%
  summarise(
    min_trip = min(trip_duration),
    mean_trip = mean(trip_duration),
    max_trip = max(trip_duration),
    median_trip = median(trip_duration),
    n = n()
  )
```

Looking at the mean for the overall trip duration which is  17.6mins, we can say that member riders has less trip duration than the overall mean trip duration. Casual riders on the other hand have a trip duration greater than the overall trip duration.   
It is also necessary to mention that whiles the duration of trip taken by casual riders on average is longer than the average trip duration of member riders, member riders take more trip than casual riders

### Total  number of trips by customer type and day of the week
```{r}
 bike_rides_df_new %>%
  select(
    customer_type,
    week_day,
    trip_duration
  ) %>%
  group_by(customer_type, week_day) %>%
  summarise(
    number_of_trips = n(),
    average_trip_mins = mean(trip_duration)
  ) %>%
  arrange(customer_type, desc(number_of_trips))
```

### Visualize total trips by customer type and day of the week
```{r}
trip_week_day<- bike_rides_df_new %>%
  select(
    customer_type,
    week_day,
    trip_duration
  ) %>%
  group_by(customer_type, week_day) %>%
  summarise(
    number_of_trips = n(),
    average_trip_mins = mean(trip_duration)
  ) %>%
  arrange(customer_type, desc(number_of_trips))

weekorder<- trip_week_day$week_day[
  order(trip_week_day$customer_type, 
        trip_week_day$number_of_trips)
  ]

trip_week_day$week_day <- factor(trip_week_day$week_day, levels = unique(weekorder))

ggplot(trip_week_day, aes(x = number_of_trips, y = week_day)) +
  geom_segment(aes(yend = week_day), xend = 0, colour = "grey50") +
  geom_point(size = 6, aes(colour = customer_type)) +
  scale_colour_manual(values = c("#00AAA0", "#FF7A5A"), limits = c("member", "casual"), guide = "none") +
  theme_bw() +
  theme(
    panel.grid.major.y = element_blank(),
    axis.title = element_text(face = "bold"),
  ) +
  labs(
    title = " Number of trip by weekday and customer type",
    x = "Number of trips",
    y = "Day of the week"
  ) +
  facet_grid(customer_type ~ ., scales = "free_y", space = "free_y")
  
```

Looking at the table and graph above, it can be concluded that, ***members*** are very active users of the bike during week days. **Members** have most of their trips on Wednesday, Tuesday and Thursday and they have less trips during the weekends. On the hand casual riders are mostly using the service during the weekends. An interesting pattern to take note is that, apart from the weekends where casual riders uses the bike service more than the member riders, member riders are consistently using the service through out the week.


### Total number of trips by customer type  and bike type
```{r}
bike_rides_df_new %>%
  select(
    bike_type,
    trip_duration,
    customer_type
  ) %>%
  group_by(bike_type, customer_type) %>%
  summarise(
    average_trip_mins = mean(trip_duration),
    number_of_trips = n()
  ) %>%
  arrange(
    customer_type,
    desc(average_trip_mins)
  )
```

### Visualize Total number of trips by bike and customer type
```{r}
bike_type_number_trips<- bike_rides_df_new %>%
  select(
    bike_type,
    trip_duration,
    customer_type
  ) %>%
  group_by(bike_type, customer_type) %>%
  summarise(
    average_trip_mins = mean(trip_duration),
    number_of_trips = n()
  ) %>%
  arrange(
    customer_type,
    (average_trip_mins)
  )

bike_type_number_trips<- bike_type_number_trips %>% 
  mutate(percentage = number_of_trips/sum(number_of_trips))
  
ggplot(
  bike_type_number_trips, aes(x = bike_type, y = number_of_trips, fill = customer_type)
) +
  geom_col(position = "dodge", color = "black", width = 0.8) +
  labs(
    title = "Total number of trips by bike type and customer type",
    x = "bike type",
    y = "number of trips",
    fill = "customer type"
  ) +
  theme(
    axis.title.x = element_text(face = "bold"),
    axis.text.x = element_text(angle = 60)
  ) +
  theme_bw() +
  scale_fill_manual(values = c("#00AAA0", "#FF7A5A"), guide = "none") +
  geom_label(aes(label = percent(percentage, scale = 100)),
    position = position_dodge(0.8),
    size = 3,
    vjust = 1.2,
    hjust = 0.5,
    colour = "black"
  ) +
  facet_wrap(~customer_type)
  
```



### Total monthly trips by customer type
```{r}
bike_rides_df_new %>%
  select(
    month,
    customer_type,
    trip_duration
  ) %>%
  group_by(month, customer_type) %>%
  summarise(
    average_trip_mins = mean(trip_duration),
    number_of_trips = n()
  ) %>%
  arrange(desc(number_of_trips)) 
```


### Top most demanded hours (24 hours) of bikes by  customer type
```{r}
hourly_demand<- bike_rides_df_new %>%
  select(
    start_time,
    week_day,
    customer_type
  ) %>% 
  group_by(customer_type, start_time, week_day) %>%
  summarise(number_of_trips = n()) %>%
  arrange(
    desc(number_of_trips)
  ) 
  
```

```{r}

  
```

