---
title: "Cyclistic bike-share analysis case study!"
author: "Ameyaw Ahmed Analysis"
date: '2022-07-15'
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Scenario
You are a junior data analyst working in the marketing analyst team at Cyclistic, a bike-share company in Chicago. The director of marketing believes the company’s future success depends on maximizing the number of annual memberships. Therefore, your team 
wants to understand how casual riders and annual members use Cyclistic bikes differently. From these insights, your team will design a new marketing strategy to convert casual riders into annual members. But first, Cyclistic executives must approve your recommendations, so they must be backed up with compelling data  insights and 
professional data visualizations.  

# Business Task
 The goal of this project is to understand how casual riders and annual members use Cyclistic bikes differently and to find insights that will help my team design a new marketing strategy to convert casual riders into annual members. Hence the **key questions** that my team and I are trying to answer are 

- *1.* How do annual members and casual riders use Cyclistic bikes differently?
- *2.* Why would casual riders buy Cyclist annual memberships?
- *3.* How can Cyclistic use digital media to influence casual riders to become members?  

 My main task is to answer the first Question by finding insigths which would help the
 marketing team answer the first question.
 
# Source of data
The data has been made available by Motivate International Inc. under this [license](https://www.divvybikes.com/data-license-agreement). The data is a monthly data which begins at July 11th, 2021 andf ends at July 14th 2021. The data can be found [here](https://divvy-tripdata.s3.amazonaws.com/index.html).

## Libraries
```{r message=FALSE, warning=FALSE, include=FALSE}
library(tidyverse)
library(janitor)
library(here)
library(lubridate)
```

## Impoting files
```{r include=FALSE}
Jul_firstyr <- read_csv(here::here("Dataset/Jul_2021_tripdata.csv"))
Aug_firstyr <- read_csv(here::here("Dataset/Aug_2021_tripdata.csv"))
Sep_firstyr <- read_csv(here::here("Dataset/Sep_2021_tripdata.csv"))
Oct_firstyr <- read.csv(here::here("Dataset/Oct_2021_tripdata.csv"))
Nov_firstyr <- read_csv(here::here("Dataset/Nov_2021_tripdata.csv"))
Dec_firstyr <- read_csv(here::here("Dataset/Dec_2021_tripdata.csv"))
Jan_secyr <- read_csv(here::here("Dataset/Jan_2022_tripdata.csv"))
Feb_secyr <- read_csv(here::here("Dataset/Feb_2022_tripdata.csv"))
Mar_secyr <- read_csv(here::here("Dataset/Mar_2022_tripdata.csv"))
Apr_secyr <- read.csv(here::here("Dataset/Apr_2022_tripdata.csv"))
May_secyr <- read_csv(here::here("Dataset/May_2022_tripdata.csv"))
Jun_secyr <- read_csv(here::here("Dataset/Jun_2022_tripdata.csv"))
Jul_secyr <- read_csv(here::here("Dataset/Jul_2022_tripdata.csv"))

```



 
 